---
title: "Scalar contradictions & gerrymandering"
subtitle: "The case of Val-de-Marne (94)"
author: "Claude Grasland"
format: 
  html:
    embed-resources: true
    echo: false
    comments: false
    warning: false
---


```{r, echo=FALSE}
library(sf, quietly = T, verbose=F,warn.conflicts = F)
library(mapsf, quietly = T, verbose=F,warn.conflicts = F)
library(dplyr, quietly = T, verbose=F,warn.conflicts = F)
library(leaflet, quietly = T, verbose=F,warn.conflicts = F)
```





The aim of this note is to examine some contradictions between the rule governing the electoral division and the current situation. We have chosen the *département* of Val-de-Marne as example because of this exceptional complexity in terms of contradictions between the different geographical divisions that should be used in the analysis

# SCALAR CONTRADICTIONS



The most fundamental units for the project are *cantons* and *cicronscription*. In the case of urbanised area like Val-de-Marne,it is not possible to build circonscription without breaking some cantons. But what is the level of discrepancy between the two divisions ?

```{r}
circ<-readRDS("data94/circ94.RDS") %>% st_transform(2154)
cant<-readRDS("data94/cant94.RDS")  %>% st_transform(2154)
com<-readRDS("data94/com94.RDS")  %>% st_transform(2154)
iris<-readRDS("data94/iris94.RDS")  %>% st_transform(2154)
grid<-readRDS("data94/grid94.RDS")  %>% st_transform(2154)
```

## Circonscriptions

```{r}
mf_map(circ, type="typo", var="circ_code", leg_pos = NA)
mf_label(circ, var="circ_code",cex = 1.2,halo = T)
mf_layout("The 11 electoral circonscription of the département Val-de-Marne",
          frame=T, scale=T, arrow=F, 
          credits= "Projet REDEC, C. Grasland, Univ. Paris Cité")
```

## Cantons

```{r}
mf_map(cant, type="typo", var="can_id",leg_pos = NA)
mf_label(cant, var="can_id",cex = 1,halo = T)
mf_layout("The 25 cantons of the département of Val-de-Marne",
          frame=T, scale=T, arrow=F, credits= "Projet REDEC, C. Grasland, Univ. Paris Cité")
```


The elaboration of 11 circonscription with 25 cantons as building block should normally be feasible by simple agregation with a minimum of cases of fragmentation of cantons. But it is not at all what is observed in the département of Val-de-Marne. 

```{r}
mf_map(circ, 
       type="typo",
       var="circ_code",
       leg_pos = NA, 
       border="white", 
       lwd=2)
mf_map(cant, type="base", col=NA, border="black", lwd=2, add=T)
mf_label(circ, var="circ_code",cex = 1.2,halo = T)
mf_layout("Differences between the limits of cantons and circonscriptions",
          frame=T, scale=T, arrow=F, credits= "Projet REDEC, C. Grasland, Univ. Paris Cité")
```

- **Comment** : The reason of the contradiction is the electoral reform of departemental election of 2014 which has divided their number by two in order to insure gender equality. Before 2014, the circonscription fullfilled the obligation to be made of aggregates of cantons. But it was not the case after the reform. 


## Communes 

In the french electoral system, the *communes* are a crucial scale of governance below the level of departement (electtion by *cantons*), the region (election by *département*) and the national assembly (election by *circonscription*). It is therefore not possible to understand the logic observe at the levels of cantons and circonscription without looking at the interactions with the level of commune. **The differents scales of the electoral system are strongly interlinked**. 


In many cases, the delineation of cantons has divided the most populated communes in subparts. Was it always justified ? And what was the political resason of such subdivisions ?

```{r}
mf_map(com, 
       type="typo",
       var="com_code",
       leg_pos = NA, 
       border="white", 
       lwd=1)
mf_map(cant, type="base", col=NA, border="black", lwd=2, add=T)
mf_label(com, var="com_code",cex = 0.5,halo = T,overlap = F)
mf_layout("Differences between the limits of cantons and communes",
          frame=T, scale=T, arrow=F, credits= "Projet REDEC, C. Grasland, Univ. Paris Cité")
```


- **Comment** : The map reveals many cases of division of communes between two cantons. This situations appears generally justified by the size of the most populated communes like Vitry-sur-Seine (94081), Créteil (94028), Champigny-sur-Marne. But it is more questionnable in the case of smaller communes like Villeneuvve-Saint-Georges (94078) or Vincennes (94080). We can suspect that some of the fragmentations of communes has been designed in order to favour the victory of one political party in some cantons. 

Admitting that it was necessary to divide some communes at the level of cantons, wasn't it possible to proceed to their reunification at the level of circonscription ? It should normally be possible but it is not always the case as we can see on the map below. 

```{r}
mf_map(com, 
       type="typo",
       var="com_code",
       leg_pos = NA, 
       border="white", 
       lwd=1)
mf_map(circ, type="base", col=NA, border="black", lwd=2, add=T)
mf_label(com, var="com_code",cex = 0.5,halo = T,overlap = F)
mf_layout("Differences between the limits of circonscriptions and communes",
          frame=T, scale=T, arrow=F, credits= "Projet REDEC, C. Grasland, Univ. Paris Cité")
```

- ***Comment*** : Some communal divisions at cantonal level has been removed at the level of circonsciption like in the case of Saint-Maur-des-Fossés (94068) or Vincennes (94080). But in other cases the divisions of communes has been maintained and eventually complexified by the apparition of new divisions of communes that do not follow the one introduced by canton. It is particularly the case for Créteil (94028), Vitry-sur-Seine (94081) and Champigny-sur-Marne (94017). This strange result is without any doubt a case of gerrymandering and it is certainly not by chance if the commmunes that has been subject to the most contradictory divisions belong to the same political orientation (left) when reunified communes be;long to the other political orientation (right).


# GERRYMANDERING

The evaluation of political strategies that has been used by the author of territorial delineation of cantons and circonscription can not be clarified by data collected at communal level in the urbanized area. To analyze social homogeneity of population in terms of social and economic characteristics or political orientations, we need informations at a lower scale. But the great problem for this task is the fact that socio-economic data and political data are not available on the same territorial mesh.

## Data collection

### Socio-econcomic data by IRIS

Since many years, the french census offers extraordinary precise data at the level of IRIS which are spatial blocks of population with a minimum of approximatively 2000 inhabitants. Such data covers all aspect of social and ecomic life and are associated to many tables available on the website of INSEE. And they can even be related to individual databases where the location of household and people can be precisely associated to the IRIS. It is therefore very simple to analyze the internal social characteristics of circonscription and produce related maps. 

### Electoral data by votation office

The votation office are places where citizens are going to make their votations. It is generally a primary school or another municipal infrastructure that play the role of office. The size is a bit smaller than IRIS (approx. 1000 citizens per office). The main problem is that the precise geography of votation office is difficult to represent on a map because it is a list of streets and adresses and not a polygon with borders. Many researchers has tried to indentify theses limits and to cross the area of votation office with the IRIS in order to compare socio-economic data and electoral choices. But it remains difficult and subject to approximations. It is nevertheless of very high interest as we can demonstate by two simple examples.

## Gerrymandering



### Case 1 : Cracking

Cracking is a method of gerrymandering where the objective is to split the opponent in area where the party of the author of gerrymandeing is expected to obtain 55-60% of voices. The voices of the opponet are therefore lost.

```{r}
seliris <- iris %>% filter(circ_code==1) 

par(mfrow=c(1,2))

seliris$pop<-seliris$pct_csp_ouvrier+seliris$pct_csp_employe
## Map 1
mf_map(seliris, 
       type="choro",
       var="pop",
       pal="Reds",
       border="white", 
         breaks=c(5, 15, 25, 35, 100),
       leg_title = "%",
       leg_val_rnd = 0,
       lwd=0.5)
mf_map(circ, type="base", col=NA, border="black", lwd=2, add=T)
mf_map(com, type="base", col=NA, border="black", lwd=1, add=T)

mf_label(com, var="com_name",cex = 0.5,halo = T,overlap = F)
mf_layout("% of low-middle class",
          frame=T, scale=T, arrow=F, credits= "Source : INSEE, Population census 2021")

seliris$right<-seliris$score_LREM+seliris$score_LR
## Map 2
mf_map(seliris, 
       type="choro",
       var="right",
       pal="Blues",
       border="white", 
       breaks=c(5, 15, 25, 35, 100),
       leg_title = "%",
       leg_val_rnd = 0,
       lwd=0.5)
mf_map(circ, type="base", col=NA, border="black", lwd=2, add=T)
mf_map(com, type="base", col=NA, border="black", lwd=1, add=T)

mf_label(com, var="com_name",cex = 0.5,halo = T,overlap = F)
mf_layout("% of vote for right",
          frame=T, scale=T, arrow=F, credits= "Source : Results of european electoon 2024")


```

- **Comment** : despite some missing values the two maps clearly help to understand the political strategy of the author of the delineation of the 1st Circonscription of val de Marne. One the one hand, we have the populated and rich commune of Saint-Maur-des-Fossés where the victory of the right-wing political party is obvious (generally with 60 to 70% of vote). On the other hand, we have neighboring communes or pieces of communes where left-wing party are the winners with a large majority. The strategy will be therefore to mix a small proportion of this left-wing area to the right-wing commune of Saint-Maur-des-Fossés in order to keep the majority (with approx. 55%) and limit the election of a left-wing deputy in other circonscriptions.


### Case 2 : Packing

Packing is the reverse method of grouping all the voices of the opponent in a single circonscription with 80-90% of voices. It is another way to vaste the voices as it would be sufficient to obtain 50% to win a circonscription.


```{r}
seliris <- iris %>% filter(circ_code==2) 

par(mfrow=c(1,2))

seliris$pop<-seliris$pct_csp_ouvrier+seliris$pct_csp_employe
## Map 1
mf_map(seliris, 
       type="choro",
       var="pop",
       pal="Reds",
       border="white", 
         breaks=c(5, 15, 25, 35, 100),
       leg_title = "%",
       leg_val_rnd = 0,
       lwd=0.5)
mf_map(circ, type="base", col=NA, border="black", lwd=2, add=T)
mf_map(com, type="base", col=NA, border="black", lwd=1, add=T)

mf_label(com, var="com_name",cex = 0.5,halo = T,overlap = F)
mf_layout("% of low-middle class",
          frame=T, scale=T, arrow=F, credits= "Source : INSEE, Population census 2021")

seliris$right<-seliris$score_LREM+seliris$score_LR
## Map 2
mf_map(seliris, 
       type="choro",
       var="right",
       pal="Blues",
       border="white", 
       breaks=c(5, 15, 25, 35, 100),
       leg_title = "%",
       leg_val_rnd = 0,
       lwd=0.5)
mf_map(circ, type="base", col=NA, border="black", lwd=2, add=T)
mf_map(com, type="base", col=NA, border="black", lwd=1, add=T)

mf_label(com, var="com_name",cex = 0.5,halo = T,overlap = F)
mf_layout("% of vote for right",
          frame=T, scale=T, arrow=F, credits= "Source : Results of european electoon 2024")


```

- ***Comment :*** The very strange form of the second circonscription of the Val-de-Marne illustrate the second strategy of gerrymandering which is to concentrate the voices of the opponent party in a limited number of circonscription. The "salamander" form of the circonscription was necessary to put together areas where blue collar and middle class are concentrated and where the right-wing party had no chance to win. Putting tohether the communes of Orly, Thiais and part of Créteil was a very efficient way to waste the voices of the left-wing party ...

