---
title: "Visualisation tool"
subtitle: "The case of Val-de-Marne (94)"
author: "Claude Grasland"
format: 
  html:
    embed-resources: true
    echo: false
    comments: false
    warning: false
---


```{r, echo=FALSE}
library(sf, quietly = T, verbose=F,warn.conflicts = F)
library(mapsf, quietly = T, verbose=F,warn.conflicts = F)
library(dplyr, quietly = T, verbose=F,warn.conflicts = F)
library(leaflet, quietly = T, verbose=F,warn.conflicts = F)
library(sp)
library(spdep)
library(adespatial)
```



```{r}
mydep <-"94"
mynomdep <-"Val-de-Marne"
capnom <- "Créteil"
capcode <- "94028"



bv1<-readRDS(paste0("map/bv",mydep,"/mapbv_voronoi.RDS")) %>% st_transform(2154)
ci1 <-readRDS(paste0("map/bv",mydep,"/mapcir_voronoi.RDS")) %>% st_transform(2154)
cap1 <- bv1 %>%filter(comcode==capcode) %>% group_by(comcode,comname) %>% summarise()
bv2<-readRDS(paste0("map/bv",mydep,"/mapbv_inscrits.RDS")) %>% st_transform(2154)
ci2 <-readRDS(paste0("map/bv",mydep,"/mapcir_inscrits.RDS")) %>% st_transform(2154)
ci2 <-st_buffer(ci2,10)
cap2 <- bv2 %>% filter(comcode==capcode)  %>% group_by(comcode, comname) %>% summarise()

```


# Area and population

We propose to visualize the delineation of circonscription with two complementary maps.

## Geometric space

We use here the classical map in geometric space with eucldean distance. We can therefore viusalize large surfaces corresponding to low density area and small surfaces corresponding to high density area.

```{r}

par(mfrow=c(1,1))

## Geographic space
bv<-bv1
ci<-ci1
ca<-cap1

  nbc<-dim(ci)[1]
  bv$Num <- substr(bv$circode,3,4)
  bv<-st_make_valid(bv)
  nbv<-dim(bv)[1]
  
  mf_map(bv, type="typo",var="Num", 
         alpha=0.2, 
         leg_pos =  "topright", leg_title = "Circ.",
         border="gray30",
         lwd=0.2)
  mf_map(ci, type="base",col=NA,lwd=2,border="black", add=T)
    mf_map(ca, type="base", col=NA,lwd=2, border="red", add=T,lty=3)
    mf_label(ca,var="comname", col="red", halo=T)

    mf_layout("Area in geometric space",
          frame=T,
          scale=T,
          arrow=F,
          credits = "Sources : INSEE, IGN, Min. Intérieur (2022-2026)
Author : Grasland C., 2026, Preparation of  ANR REDEC")
   

    
```

## Demographic space

We use here a cartogramme transformation of voting office in order to produce a demographic space where the area is proportional to the number of people officially allowed to vote. The distance between two points is the number of inhabitant between origin and destination (the Stouffer distance).

```{r}
## Geographic space
bv<-bv2
ci<-ci2
ca<-cap2

  ci <-st_buffer(ci,10)
  nbc<-dim(ci)[1]
  bv$Num <- substr(bv$circode,3,4)
  bv<-st_make_valid(bv)
  nbv<-dim(bv)[1]
  
  mf_map(bv, type="typo",var="Num", 
         alpha=0.2, 
         leg_pos = "topright", leg_title = "Circ.",
         border="gray30",
         lwd=0.2)
  mf_map(ci, type="base",col=NA,lwd=2,border="black", add=T)
  
    mf_map(ca, type="base", col=NA,lwd=2, border="red", add=T,lty=3)
    mf_label(ca,var="comname", col="red", halo=T)

    mf_layout("Population in demographic space",
          frame=T,
          scale=F,
          arrow=F,
          credits = "Sources : INSEE, IGN, Min. Intérieur (2022-2026)
Author : Grasland C., 2026, Preparation of ANR REDEC") 
    
```

# Political orientation

The first turn of the presidential election of 2022 is characterized by the emergence of three blocks, represented by the candidates E. Macron (center right), J.L. Melenchon (left) and Marine Le Pen (far right).  It offers a good view of the political landscape before the parliament elections that follows.



```{r}
mapchoro <-function(mybv=bv,
                    myci=ci, 
                    myvar="Mac", 
                    myname="Macron",
                    mypal="Oranges",
                    mycred = "Author : Grasland C., 2026, Preparation of  ANR REDEC",
                    myscale = T){
  
 bv<-mybv
 ci<-myci
 bv$x <-bv[[myvar]]
bv$x<-100*bv$x/bv$exp
  mf_map(bv, type="choro",
         var="x", 
         breaks="q6",
         pal=mypal,
         alpha=0.6,
         leg_pos =  "topright", 
         leg_title = "% expr",
                  leg_val_rnd = 0,
         border="gray30",
         lwd=0.1)
  mf_map(ci, type="base",col=NA,lwd=1,border="black", add=T)

    mf_layout(myname,
          frame=T,
          scale=myscale,
          arrow=F,
          credits = mycred)
  
}

   
    
```


## In geometric space

What is the distribution of choices by voting office and criconscription in geometric space ?

```{r, fig.with=6, fig.height=15}

par(mfrow=c(3,1))

mapchoro(bv1,ci1,"Mac","Macron","Oranges")
mapchoro(bv1,ci1,"Mel","Mélenchon","Reds")
mapchoro(bv1,ci1,"Lep","Le Pen","Blues")


    
```




## In demographic space

What is the picture if we transform the map in a cartogramme where the area is proportional to the number of registered people that was likely to vote.

```{r, fig.with=6, fig.height=12}
par(mfrow=c(3,1))

mapchoro(bv2,ci2,"Mac","Macron","Oranges")
mapchoro(bv2,ci2,"Mel","Mélenchon","Reds")
mapchoro(bv2,ci2,"Lep","Le Pen","Blues")
```

# Partisan bias ?

To what extent can be observe a partisan bias in the delineation of circonscription ? Do we observe specific concentration of votes in favour of one of the polical block in selected circonscriptions ? Can we suspect the existence of gerrymandering i.e. an intention to favour some political groups by methods of packing and cracking ?

The delineation of circonscription are inherited from an historical period where the main political division was an opposition between left and right. But with the existence of three poles (illustrated by the results of the vote for Macron, Mélenchon and Le Pen), the bias has been strongly modified.


```{r}
mapnetwork <- function(mapbv = bv1,
                       mapcir = ci1,
                       mapca = cap1,
                       cirnum = "3801",
                       mytitle = "Geometric space",
                       myscale = F,
                      mycred = "",
                      leg = TRUE
                       ) {
  
  
 #mapbv <- bv1
 #mapcir <- ci1
# cirnum <- "3801" 
  
bv<-mapbv
cir <-mapcir
ca <-mapca

 mapdon <- bv %>% filter(circode==cirnum)
 

# Calcul de la matrice de contiguïté
mat_conti <- st_intersects(mapdon, mapdon, sparse = FALSE)
colnames(mat_conti) <- mapdon$bvocode
rownames(mat_conti) <- mapdon$bvocode

# Suppression de la moitié de la matrice (et de la diagonale)
mat_conti[lower.tri(mat_conti, diag = TRUE)] <- FALSE

# Construction d'un tableau de lien (i, j) de contiguïté
reg_link_contig <- as.data.frame.table(mat_conti, responseName = "contig") |> 
  filter(contig == TRUE)

# Création de la couche géographique de liens
reg_links_contig  <- mf_get_links(x = mapdon, 
                                  df = reg_link_contig, 
                                  x_id = "bvocode", 
                                  df_id = c("Var1","Var2"))

# Graphe de voisinage
sel <- bv %>% filter(comcode %in% mapdon$comcode) %>% group_by(comname)%>%
  summarise() %>% st_buffer(10)

mf_map(mapdon, type="base", col="gray90", border=NA)
#mf_map(sel, type="base", var="comcode", add=T, col=NA, lwd=2)

mf_map(reg_links_contig , col = "black", add = TRUE, lwd=0.4)
#mf_label(sel, var = "comname", cex = 0.8, col = "blue3", halo = TRUE, bg = "white")
if (leg==T) {
mf_prop_typo(mapdon, var=c("ins","top"), inches=0.07, leg_title = c("nb. inscrits", "Pres. elect. 2022"))
} else {
mf_prop_typo(mapdon, var=c("ins","top"), inches=0.07, leg_pos=c(NA,NA))  
  
}
mf_map(ca, type="base",col=NA,border="black", add=T, lwd=2, lty=3)
#mf_label(ca, var = capnom, col="blue", halo=T)
mf_layout(title = paste(cirnum, ":", mytitle), credits = "", arrow=F, frame=T, scale=myscale)


}




```




```{r, fig.height=35, fig.width=8}

listcode<-ci$circode
nb<-length(listcode)
par(mfrow=c(nb,2))
for(i in 1:nb) {
mapnetwork(mapbv = bv1,mapcir = ci1,mapca = cap1,cirnum = listcode[i], "Geometric space", leg=T)
mapnetwork(mapbv = bv2,mapcir = ci2, mapca = cap2,cirnum = listcode[i], "Demographic space", leg=F)
}


```




