---
title: "Partisan bias & Gerrymandering "
subtitle: "The case of Isère (38)"
author: "Claude Grasland"
format: 
  html:
    embed-resources: true
    echo: false
    comments: false
    warning: false
---


```{r, echo=FALSE}
library(sf, quietly = T, verbose=F,warn.conflicts = F)
library(mapsf, quietly = T, verbose=F,warn.conflicts = F)
library(dplyr, quietly = T, verbose=F,warn.conflicts = F)
library(leaflet, quietly = T, verbose=F,warn.conflicts = F)
library(sp)
library(spdep)
library(adespatial)
```



The aim of this note is to examine the problem of delineation of circonscription in the département of Isère (38) characterized by strong geographical constraint, unequal density and gread diversity of political orientations. We will focus in particular in the analysis of the southern part of the département around the agglomeration of Grenoble and the mountain valleys around it.

# DELINEATION OF CIRCONSCRIPTION

## Geometric space

```{r}
mydep <- "38"
mynomdep <-"Isère"



bv1<-readRDS(paste0("map/bv",mydep,"/mapbv_voronoi.RDS")) %>% st_transform(2154)
ci1 <-readRDS(paste0("map/bv",mydep,"/mapcir_voronoi.RDS")) %>% st_transform(2154)
bv2<-readRDS(paste0("map/bv",mydep,"/mapbv_inscrits.RDS")) %>% st_transform(2154)
ci2 <-readRDS(paste0("map/bv",mydep,"/mapcir_inscrits.RDS")) %>% st_transform(2154)
  ci2 <-st_buffer(ci2,10)

par(mfrow=c(1,1))

## Geographic space
bv<-bv1
ci<-ci1

  nbc<-dim(ci)[1]
  bv$Num <- substr(bv$circode,3,4)
  bv<-st_make_valid(bv)
  nbv<-dim(bv)[1]
  
  mf_map(bv, type="typo",var="Num", 
         alpha=0.2, 
         leg_pos =  "topright", leg_title = "Circ.",
         border="gray30",
         lwd=0.2)
  mf_map(ci, type="base",col=NA,lwd=2,border="black", add=T)

    mf_layout("Area in geometric space",
          frame=T,
          scale=T,
          arrow=F,
          credits = "Sources : INSEE, IGN, Min. Intérieur (2022-2026)
Author : Grasland C., 2026, Preparation of  ANR REDEC")
   

    
```

## Demographic space


```{r}
## Geographic space
bv<-bv2
ci<-ci2

  ci <-st_buffer(ci,10)
  nbc<-dim(ci)[1]
  bv$Num <- substr(bv$circode,3,4)
  bv<-st_make_valid(bv)
  nbv<-dim(bv)[1]
  
  mf_map(bv, type="typo",var="Num", 
         alpha=0.2, 
         leg_pos = "topright", leg_title = "Circ.",
         border="gray30",
         lwd=0.2)
  mf_map(ci, type="base",col=NA,lwd=2,border="black", add=T)
  
  

    mf_layout("Population in demographic space",
          frame=T,
          scale=F,
          arrow=F,
          credits = "Sources : INSEE, IGN, Min. Intérieur (2022-2026)
Author : Grasland C., 2026, Preparation of ANR REDEC") 
    
```

# PRESIDENTIAL ELECTION 2022 (first turn)

The first turn of the presidential election of 2022 is characterized by the emergence of three blocks, represented by the candidates E. Macron (center right), J.L. Melenchon (left) and Marine Le Pen (far right).  It offers a good view of the political landscape before the parliament elections that follows.

## In geometric space

What is the distribution of choices by voting office and criconscription in geometric space ?

```{r, fig.with=6, fig.height=15}
bv<-bv1
ci<-ci1
par(mfrow=c(4,1))

## Geographic space



bv$x<-100*bv$Mac/bv$exp
  mf_map(bv, type="choro",var="x", 
         breaks="q6",
         pal="Oranges",
         alpha=0.6,
         leg_pos =  "topright", 
         leg_title = "% expr",
                  leg_val_rnd = 0,
         border="gray30",
         lwd=0.1)
  mf_map(ci, type="base",col=NA,lwd=1,border="black", add=T)

    mf_layout("Macron",
          frame=T,
          scale=T,
          arrow=F,
          credits = "Author : Grasland C., 2026, Preparation of  ANR REDEC")   
    
    
bv$x<-100*bv$Mel/bv$exp
  mf_map(bv, type="choro",var="x", 
         breaks="q6",
         pal="Reds",
         alpha=0.6,
         leg_pos =  "topright", 
         leg_title = "% expr.",
                  leg_val_rnd = 0,
         border="gray30",
         lwd=0.1)
  mf_map(ci, type="base",col=NA,lwd=1,border="black", add=T)

    mf_layout("Melenchon",
          frame=T,
          scale=T,
          arrow=F,
          credits = "Author : Grasland C., 2026, Preparation of  ANR REDEC")   
    
    
    
bv$x<-100*bv$Lep/bv$exp
  mf_map(bv, type="choro",var="x", 
         breaks="q6",
         pal="Blues",
         alpha=0.6,
         leg_pos =  "topright", 
         leg_title = "% expr.",
                  leg_val_rnd = 0,
         border="gray30",
         lwd=0.1)
  mf_map(ci, type="base",col=NA,lwd=1,border="black", add=T)

    mf_layout("Le Pen",
          frame=T,
          scale=T,
          arrow=F,
          credits = "Author : Grasland C., 2026, Preparation of  ANR REDEC")    
    
    
bv$x<-100*(bv$exp - bv$Lep -bv$Mac - bv$Mel)/bv$exp
  mf_map(bv, type="choro",var="x", 
         breaks="q6",
         pal="Grays",
         alpha=0.6,
         leg_pos =  "topright", 
         leg_title = "% expr.",
         leg_val_rnd = 0,
         border="gray30",
         lwd=0.1)
  mf_map(ci, type="base",col=NA,lwd=1,border="black", add=T)

    mf_layout("Other candidates",
          frame=T,
          scale=T,
          arrow=F,
          credits = "Author : Grasland C., 2026, Preparation of  ANR REDEC")     
   

    
```




## In demographic space

What is the picture if we transform the map in a cartogramme where the area is proportional to the number of registered people that was likely to vote.

```{r, fig.with=6, fig.height=15}
bv<-bv2
ci<-ci2
par(mfrow=c(4,1))

## Geographic space



bv$x<-100*bv$Mac/bv$exp
  mf_map(bv, type="choro",var="x", 
         breaks="q6",
         pal="Oranges",
         alpha=0.6,
         leg_pos =  "topright", 
         leg_title = "% expr",
                  leg_val_rnd = 0,
         border="gray30",
         lwd=0.1)
  mf_map(ci, type="base",col=NA,lwd=1,border="black", add=T)

    mf_layout("Macron",
          frame=T,
          scale=T,
          arrow=F,
          credits = "Author : Grasland C., 2026, Preparation of  ANR REDEC")   
    
    
bv$x<-100*bv$Mel/bv$exp
  mf_map(bv, type="choro",var="x", 
         breaks="q6",
         pal="Reds",
         alpha=0.6,
         leg_pos =  "topright", 
         leg_title = "% expr.",
                  leg_val_rnd = 0,
         border="gray30",
         lwd=0.1)
  mf_map(ci, type="base",col=NA,lwd=1,border="black", add=T)

    mf_layout("Melenchon",
          frame=T,
          scale=T,
          arrow=F,
          credits = "Author : Grasland C., 2026, Preparation of  ANR REDEC")   
    
    
    
bv$x<-100*bv$Lep/bv$exp
  mf_map(bv, type="choro",var="x", 
         breaks="q6",
         pal="Blues",
         alpha=0.6,
         leg_pos =  "topright", 
         leg_title = "% expr.",
                  leg_val_rnd = 0,
         border="gray30",
         lwd=0.1)
  mf_map(ci, type="base",col=NA,lwd=1,border="black", add=T)

    mf_layout("Le Pen",
          frame=T,
          scale=T,
          arrow=F,
          credits = "Author : Grasland C., 2026, Preparation of  ANR REDEC")    
    
    
bv$x<-100*(bv$exp - bv$Lep -bv$Mac - bv$Mel)/bv$exp
  mf_map(bv, type="choro",var="x", 
         breaks="q6",
         pal="Grays",
         alpha=0.6,
         leg_pos =  "topright", 
         leg_title = "% expr.",
         leg_val_rnd = 0,
         border="gray30",
         lwd=0.1)
  mf_map(ci, type="base",col=NA,lwd=1,border="black", add=T)

    mf_layout("Other candidates",
          frame=T,
          scale=T,
          arrow=F,
          credits = "Author : Grasland C., 2026, Preparation of  ANR REDEC")     
   

    
```

# GERRYMANDERING

We propose to appreciate the effect of gerrymandering and partisan bias on selected circonscription characterized either by a strange geometric form and/or by a strong  clustering of voices for one or several candidates arrived in first places at the presidential election of 2022. But we have to keep in mind that the clustering was not necessary the result of a "conspiration". But rather the inheritage of a long political history. 

## Circonscription n°1

```{r}

par(mfrow=c(1,2))

bv<-bv1

mapdon <- bv %>% filter(circode=="3801")

# Calcul de la matrice de contiguïté
mat_conti <- st_intersects(mapdon, mapdon, sparse = FALSE)
colnames(mat_conti) <- mapdon$bvocode
rownames(mat_conti) <- mapdon$bvocode

# Suppression de la moitié de la matrice (et de la diagonale)
mat_conti[lower.tri(mat_conti, diag = TRUE)] <- FALSE

# Construction d'un tableau de lien (i, j) de contiguïté
reg_link_contig <- as.data.frame.table(mat_conti, responseName = "contig") |> 
  filter(contig == TRUE)

# Création de la couche géographique de liens
reg_links_contig  <- mf_get_links(x = mapdon, 
                                  df = reg_link_contig, 
                                  x_id = "bvocode", 
                                  df_id = c("Var1","Var2"))




# Graphe de voisinage
sel <- bv %>% filter(comcode %in% mapdon$comcode) %>% group_by(comname)%>%
  summarise() %>% st_buffer(10)

mf_map(mapdon, type="base", col="gray90", border=NA)
#mf_map(sel, type="base", var="comcode", add=T, col=NA, lwd=2)

mf_map(reg_links_contig , col = "black", add = TRUE, lwd=0.4)
#mf_label(sel, var = "comname", cex = 0.8, col = "blue3", halo = TRUE, bg = "white")
mf_prop_typo(mapdon, var=c("ins","top"), inches=0.07, leg_title = c("nb. inscrits", "Pres. elect. 2022"))
#mf_map(Amcont, type="bv",col=NA,border="red", add=T)
mf_layout("Geometric space", credits = "", arrow=F, frame=T)


bv<-bv2

mapdon <- bv %>% filter(circode=="3801")

# Calcul de la matrice de contiguïté
mat_conti <- st_intersects(mapdon, mapdon, sparse = FALSE)
colnames(mat_conti) <- mapdon$bvocode
rownames(mat_conti) <- mapdon$bvocode

# Suppression de la moitié de la matrice (et de la diagonale)
mat_conti[lower.tri(mat_conti, diag = TRUE)] <- FALSE

# Construction d'un tableau de lien (i, j) de contiguïté
reg_link_contig <- as.data.frame.table(mat_conti, responseName = "contig") |> 
  filter(contig == TRUE)

# Création de la couche géographique de liens
reg_links_contig  <- mf_get_links(x = mapdon, 
                                  df = reg_link_contig, 
                                  x_id = "bvocode", 
                                  df_id = c("Var1","Var2"))




# Graphe de voisinage
sel <- bv %>% filter(comcode %in% mapdon$comcode) %>% group_by(comname)%>%
  summarise() %>% st_buffer(10)

mf_map(mapdon, type="base", col="gray90", border=NA)
#mf_map(sel, type="base", var="comcode", add=T, col=NA, lwd=2)

mf_map(reg_links_contig , col = "black", add = TRUE, lwd=0.4)
#mf_label(sel, var = "comname", cex = 0.8, col = "blue3", halo = TRUE, bg = "white")
mf_prop_typo(mapdon, var=c("ins","top"), inches=0.07, leg_title = c("nb. inscrits", "Pres. elect. 2022"),leg_pos = c(NA,NA))
#mf_map(Amcont, type="bv",col=NA,border="red", add=T)
mf_layout("Demographic space", credits = "", arrow=F, frame=T, scale=F)

```

## Circonscription n°2

```{r}

par(mfrow=c(1,2))

bv<-bv1

mapdon <- bv %>% filter(circode=="3802")

# Calcul de la matrice de contiguïté
mat_conti <- st_intersects(mapdon, mapdon, sparse = FALSE)
colnames(mat_conti) <- mapdon$bvocode
rownames(mat_conti) <- mapdon$bvocode

# Suppression de la moitié de la matrice (et de la diagonale)
mat_conti[lower.tri(mat_conti, diag = TRUE)] <- FALSE

# Construction d'un tableau de lien (i, j) de contiguïté
reg_link_contig <- as.data.frame.table(mat_conti, responseName = "contig") |> 
  filter(contig == TRUE)

# Création de la couche géographique de liens
reg_links_contig  <- mf_get_links(x = mapdon, 
                                  df = reg_link_contig, 
                                  x_id = "bvocode", 
                                  df_id = c("Var1","Var2"))




# Graphe de voisinage
sel <- bv %>% filter(comcode %in% mapdon$comcode) %>% group_by(comname)%>%
  summarise() %>% st_buffer(10)

mf_map(mapdon, type="base", col="gray90", border=NA)
#mf_map(sel, type="base", var="comcode", add=T, col=NA, lwd=2)

mf_map(reg_links_contig , col = "black", add = TRUE, lwd=0.4)
#mf_label(sel, var = "comname", cex = 0.8, col = "blue3", halo = TRUE, bg = "white")
mf_prop_typo(mapdon, var=c("ins","top"), inches=0.07, leg_title = c("nb. inscrits", "Pres. elect. 2022"))
#mf_map(Amcont, type="bv",col=NA,border="red", add=T)
mf_layout("Geometric space", credits = "", arrow=F, frame=T)


bv<-bv2

mapdon <- bv %>% filter(circode=="3802")

# Calcul de la matrice de contiguïté
mat_conti <- st_intersects(mapdon, mapdon, sparse = FALSE)
colnames(mat_conti) <- mapdon$bvocode
rownames(mat_conti) <- mapdon$bvocode

# Suppression de la moitié de la matrice (et de la diagonale)
mat_conti[lower.tri(mat_conti, diag = TRUE)] <- FALSE

# Construction d'un tableau de lien (i, j) de contiguïté
reg_link_contig <- as.data.frame.table(mat_conti, responseName = "contig") |> 
  filter(contig == TRUE)

# Création de la couche géographique de liens
reg_links_contig  <- mf_get_links(x = mapdon, 
                                  df = reg_link_contig, 
                                  x_id = "bvocode", 
                                  df_id = c("Var1","Var2"))




# Graphe de voisinage
sel <- bv %>% filter(comcode %in% mapdon$comcode) %>% group_by(comname)%>%
  summarise() %>% st_buffer(10)

mf_map(mapdon, type="base", col="gray90", border=NA)
#mf_map(sel, type="base", var="comcode", add=T, col=NA, lwd=2)

mf_map(reg_links_contig , col = "black", add = TRUE, lwd=0.4)
#mf_label(sel, var = "comname", cex = 0.8, col = "blue3", halo = TRUE, bg = "white")
mf_prop_typo(mapdon, var=c("ins","top"), inches=0.07, leg_title = c("nb. inscrits", "Pres. elect. 2022"),leg_pos = c(NA,NA))
#mf_map(Amcont, type="bv",col=NA,border="red", add=T)
mf_layout("Demographic space", credits = "", arrow=F, frame=T, scale=F)

```


## Circonscription n°3

```{r}

par(mfrow=c(1,2))

bv<-bv1

mapdon <- bv %>% filter(circode=="3803")

# Calcul de la matrice de contiguïté
mat_conti <- st_intersects(mapdon, mapdon, sparse = FALSE)
colnames(mat_conti) <- mapdon$bvocode
rownames(mat_conti) <- mapdon$bvocode

# Suppression de la moitié de la matrice (et de la diagonale)
mat_conti[lower.tri(mat_conti, diag = TRUE)] <- FALSE

# Construction d'un tableau de lien (i, j) de contiguïté
reg_link_contig <- as.data.frame.table(mat_conti, responseName = "contig") |> 
  filter(contig == TRUE)

# Création de la couche géographique de liens
reg_links_contig  <- mf_get_links(x = mapdon, 
                                  df = reg_link_contig, 
                                  x_id = "bvocode", 
                                  df_id = c("Var1","Var2"))




# Graphe de voisinage
sel <- bv %>% filter(comcode %in% mapdon$comcode) %>% group_by(comname)%>%
  summarise() %>% st_buffer(10)

mf_map(mapdon, type="base", col="gray90", border=NA)
#mf_map(sel, type="base", var="comcode", add=T, col=NA, lwd=2)

mf_map(reg_links_contig , col = "black", add = TRUE, lwd=0.4)
#mf_label(sel, var = "comname", cex = 0.8, col = "blue3", halo = TRUE, bg = "white")
mf_prop_typo(mapdon, var=c("ins","top"), inches=0.07, leg_title = c("nb. inscrits", "Pres. elect. 2022"))
#mf_map(Amcont, type="bv",col=NA,border="red", add=T)
mf_layout("Geometric space", credits = "", arrow=F, frame=T)


bv<-bv2

mapdon <- bv %>% filter(circode=="3803")

# Calcul de la matrice de contiguïté
mat_conti <- st_intersects(mapdon, mapdon, sparse = FALSE)
colnames(mat_conti) <- mapdon$bvocode
rownames(mat_conti) <- mapdon$bvocode

# Suppression de la moitié de la matrice (et de la diagonale)
mat_conti[lower.tri(mat_conti, diag = TRUE)] <- FALSE

# Construction d'un tableau de lien (i, j) de contiguïté
reg_link_contig <- as.data.frame.table(mat_conti, responseName = "contig") |> 
  filter(contig == TRUE)

# Création de la couche géographique de liens
reg_links_contig  <- mf_get_links(x = mapdon, 
                                  df = reg_link_contig, 
                                  x_id = "bvocode", 
                                  df_id = c("Var1","Var2"))




# Graphe de voisinage
sel <- bv %>% filter(comcode %in% mapdon$comcode) %>% group_by(comname)%>%
  summarise() %>% st_buffer(10)

mf_map(mapdon, type="base", col="gray90", border=NA)
#mf_map(sel, type="base", var="comcode", add=T, col=NA, lwd=2)

mf_map(reg_links_contig , col = "black", add = TRUE, lwd=0.4)

#mf_label(sel, var = "comname", cex = 0.8, col = "blue3", halo = TRUE, bg = "white")
mf_prop_typo(mapdon, var=c("ins","top"), inches=0.07, leg_title = c("nb. inscrits", "Pres. elect. 2022"),leg_pos = c(NA,NA))

#mf_map(Amcont, type="bv",col=NA,border="red", add=T)
mf_layout("Demographic space", credits = "", arrow=F, frame=T, scale=F)

```



## Circonscription n°4

```{r}

par(mfrow=c(1,2))

bv<-bv1

mapdon <- bv %>% filter(circode=="3804")

# Calcul de la matrice de contiguïté
mat_conti <- st_intersects(mapdon, mapdon, sparse = FALSE)
colnames(mat_conti) <- mapdon$bvocode
rownames(mat_conti) <- mapdon$bvocode

# Suppression de la moitié de la matrice (et de la diagonale)
mat_conti[lower.tri(mat_conti, diag = TRUE)] <- FALSE

# Construction d'un tableau de lien (i, j) de contiguïté
reg_link_contig <- as.data.frame.table(mat_conti, responseName = "contig") |> 
  filter(contig == TRUE)

# Création de la couche géographique de liens
reg_links_contig  <- mf_get_links(x = mapdon, 
                                  df = reg_link_contig, 
                                  x_id = "bvocode", 
                                  df_id = c("Var1","Var2"))




# Graphe de voisinage
sel <- bv %>% filter(comcode %in% mapdon$comcode) %>% group_by(comname)%>%
  summarise() %>% st_buffer(10)

mf_map(mapdon, type="base", col="gray90", border=NA)
#mf_map(sel, type="base", var="comcode", add=T, col=NA, lwd=2)

mf_map(reg_links_contig , col = "black", add = TRUE, lwd=0.4)
#mf_label(sel, var = "comname", cex = 0.8, col = "blue3", halo = TRUE, bg = "white")
mf_prop_typo(mapdon, var=c("ins","top"), inches=0.07, leg_title = c("nb. inscrits", "Pres. elect. 2022"))
#mf_map(Amcont, type="bv",col=NA,border="red", add=T)
mf_layout("Geometric space", credits = "", arrow=F, frame=T)


bv<-bv2

mapdon <- bv %>% filter(circode=="3804")

# Calcul de la matrice de contiguïté
mat_conti <- st_intersects(mapdon, mapdon, sparse = FALSE)
colnames(mat_conti) <- mapdon$bvocode
rownames(mat_conti) <- mapdon$bvocode

# Suppression de la moitié de la matrice (et de la diagonale)
mat_conti[lower.tri(mat_conti, diag = TRUE)] <- FALSE

# Construction d'un tableau de lien (i, j) de contiguïté
reg_link_contig <- as.data.frame.table(mat_conti, responseName = "contig") |> 
  filter(contig == TRUE)

# Création de la couche géographique de liens
reg_links_contig  <- mf_get_links(x = mapdon, 
                                  df = reg_link_contig, 
                                  x_id = "bvocode", 
                                  df_id = c("Var1","Var2"))




# Graphe de voisinage
sel <- bv %>% filter(comcode %in% mapdon$comcode) %>% group_by(comname)%>%
  summarise() %>% st_buffer(10)

mf_map(mapdon, type="base", col="gray90", border=NA)
#mf_map(sel, type="base", var="comcode", add=T, col=NA, lwd=2)

mf_map(reg_links_contig , col = "black", add = TRUE, lwd=0.4)

#mf_label(sel, var = "comname", cex = 0.8, col = "blue3", halo = TRUE, bg = "white")
mf_prop_typo(mapdon, var=c("ins","top"), inches=0.07, leg_title = c("nb. inscrits", "Pres. elect. 2022"),leg_pos = c(NA,NA))

#mf_map(Amcont, type="bv",col=NA,border="red", add=T)
mf_layout("Demographic space", credits = "", arrow=F, frame=T, scale=F)

```

## Circonscription n°5

```{r}

par(mfrow=c(1,2))

bv<-bv1

mapdon <- bv %>% filter(circode=="3805")

# Calcul de la matrice de contiguïté
mat_conti <- st_intersects(mapdon, mapdon, sparse = FALSE)
colnames(mat_conti) <- mapdon$bvocode
rownames(mat_conti) <- mapdon$bvocode

# Suppression de la moitié de la matrice (et de la diagonale)
mat_conti[lower.tri(mat_conti, diag = TRUE)] <- FALSE

# Construction d'un tableau de lien (i, j) de contiguïté
reg_link_contig <- as.data.frame.table(mat_conti, responseName = "contig") |> 
  filter(contig == TRUE)

# Création de la couche géographique de liens
reg_links_contig  <- mf_get_links(x = mapdon, 
                                  df = reg_link_contig, 
                                  x_id = "bvocode", 
                                  df_id = c("Var1","Var2"))




# Graphe de voisinage
sel <- bv %>% filter(comcode %in% mapdon$comcode) %>% group_by(comname)%>%
  summarise() %>% st_buffer(10)

mf_map(mapdon, type="base", col="gray90", border=NA)
#mf_map(sel, type="base", var="comcode", add=T, col=NA, lwd=2)

mf_map(reg_links_contig , col = "black", add = TRUE, lwd=0.4)
#mf_label(sel, var = "comname", cex = 0.8, col = "blue3", halo = TRUE, bg = "white")
mf_prop_typo(mapdon, var=c("ins","top"), inches=0.07, leg_title = c("nb. inscrits", "Pres. elect. 2022"))
#mf_map(Amcont, type="bv",col=NA,border="red", add=T)
mf_layout("Geometric space", credits = "", arrow=F, frame=T)


bv<-bv2

mapdon <- bv %>% filter(circode=="3805")

# Calcul de la matrice de contiguïté
mat_conti <- st_intersects(mapdon, mapdon, sparse = FALSE)
colnames(mat_conti) <- mapdon$bvocode
rownames(mat_conti) <- mapdon$bvocode

# Suppression de la moitié de la matrice (et de la diagonale)
mat_conti[lower.tri(mat_conti, diag = TRUE)] <- FALSE

# Construction d'un tableau de lien (i, j) de contiguïté
reg_link_contig <- as.data.frame.table(mat_conti, responseName = "contig") |> 
  filter(contig == TRUE)

# Création de la couche géographique de liens
reg_links_contig  <- mf_get_links(x = mapdon, 
                                  df = reg_link_contig, 
                                  x_id = "bvocode", 
                                  df_id = c("Var1","Var2"))




# Graphe de voisinage
sel <- bv %>% filter(comcode %in% mapdon$comcode) %>% group_by(comname)%>%
  summarise() %>% st_buffer(10)

mf_map(mapdon, type="base", col="gray90", border=NA)
#mf_map(sel, type="base", var="comcode", add=T, col=NA, lwd=2)

mf_map(reg_links_contig , col = "black", add = TRUE, lwd=0.4)

#mf_label(sel, var = "comname", cex = 0.8, col = "blue3", halo = TRUE, bg = "white")
mf_prop_typo(mapdon, var=c("ins","top"), inches=0.07, leg_title = c("nb. inscrits", "Pres. elect. 2022"),leg_pos = c(NA,NA))

#mf_map(Amcont, type="bv",col=NA,border="red", add=T)
mf_layout("Demographic space", credits = "", arrow=F, frame=T, scale=F)

```
